Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-06 paper/paper.tex
removed (beautiful) from abstract

\author[0000-0001-5725-9329]{Lauren Anderson}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, room 1005, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

@@

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.4$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise (and beautiful) because it is a generative model fit to many noisy
parallax measurements under an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an Empirical Bayes approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The prior is combined with a \tgas\ likelihood to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a parallax posterior pdf for each of 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}

>>>>

\author[0000-0001-5725-9329]{Lauren Anderson}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, room 1005, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

@@

\begin{abstract}\noindent % trust me
Converting a noisy parallax measurement into a posterior
belief over distance requires inference with a prior.
Usually this prior represents beliefs about the stellar density distribution of
the Milky Way.
However, multi-band photometry exists for a large fraction of the
\gaia\ \tgas\ Catalog and is incredibly informative about stellar distances.
Here we use \tmass\ colors for $1.4$ million \tgas\ stars to build a
noise-deconvolved empirical prior distribution for stars in color--magnitude
space.
This model contains no knowledge of stellar astrophysics or the
Milky Way, but is precise because it is a generative model fit to many noisy
parallax measurements under an assumption of stationarity.
We use the Extreme Deconvolution (\xd) algorithm---an Empirical Bayes approximation to a full
hierarchical model of the true parallax and photometry of every star---to
construct this prior.
The prior is combined with a \tgas\ likelihood to infer a precise photometric parallax estimate
and uncertainty (and full posterior) for every star.
Our parallax estimates are more precise than the \tgas\ catalog entries by a
median factor of 1.2 (14\% are more precise by a factor $>2$) and are more
precise than previous Bayesian distance estimates that use spatial priors.
We validate our parallax inferences using members of the Milky Way
star cluster M67, which is not visible as a cluster in the \tgas\ parallax
estimates, but appears as a cluster in our posterior parallax estimates.
Our results, including a parallax posterior pdf for each of 1.4
million \tgas\ stars, are available in companion electronic tables.
\end{abstract}
